Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2022-02-09 paper/paper.tex
Wednesday WIP

% Use the article doc class, with an 11 pt basic font size
\documentclass[11pt, a4paper]{article}

@@

Each retinal ganglion cell (RGC) projects growth cones (referred to
as \emph{branches}) which carry a set of $m$ receptors, $r_i$ indexed by $i$ and
expressed at levels determined by the cell soma's location on the
retinal grid. ($m$ ligands, $l_i$, are also expressed by RGCs, but these are not
used by the chemoaffinity simulation.) The expression of each receptor
varies only with respect to one spatial dimension.

The minimum number of gradients required to provide a coordinate system on the
surface of the tectum within which axons could find a target location is
$m=2$, arranged in orthogonal directions. However if there are only two
gradients, imagining that one guides axons laterally and the other guides
axons in the caudal direction, then it is necessary for there to be an
additional `stopping mechanism' to prevent all of the axons eventually being
guided to the most lateral/caudal position. To avoid the need for a different
stopping mechanism, we chose to model \emph{opposing pairs of orthogonal
gradients} with $m=4$. With opposing gradients, an axon which interacts most
strongly with a gradient in the caudal direction would be drawn to the caudal
side of the tectum, whereas an axon whose strongest interaction was to climb a
gradient in the rostral direction would be drawn to the rostral side. There is
evidence that multiple gene expression gradients are found in the brain, with
examples of opposing pairs including XX and YY.

@@

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression
as \citet{simpson_simple_2011} which is
\begin{equation} \label{e:retrcpt}
r_i(x,y) = 1.05 + 0.26 \exp(2.3 u)
\end{equation}
Where $u$ gives the direction with which the expression increases and is
substituted by $1-x$ for $r_0$, $1-y$ for $r_1$, $x$ for $r_2$ and by $y$ for
$r_3$ (Fig.\,\ref{1}B).

The tectum expresses ligands, $L_i$, for the retinal receptors, also in
orthogonal pairs of gradients. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for the form of tectal ligand expression is more ambiguous than that for
retinal receptor expression (Fig.\,\ref{f:expressions} summarises results from
the literature). We kept in mind the
possibility that tectal ligand expression may be modelled by exponential,
linear or logarithmic functions, but initially, we set it to the exponential function
\begin{equation} \label{e:tecligexp}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(2.3 u)
\end{equation}
and use the substitutions $y$, $x$, $(1-y)$ or $1-x$ in place of $u$ to obtain
the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$
(Fig.\,\ref{1}D).

@@

\color{black}
Each discrete, square element in Fig.\,\ref{f:ex}B projects one RGC axon with 4
branches to the tectum, carrying the receptor expression at the retinal
element's location.
At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the branch movement vector, is the sum of a chemoaffinity
effect and a `border effect', $\mathbf{B}_b$, which acts to retain branches
within the tectal region:
%
\begin{equation} \label{e:mv}
\mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + 0.5 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$.

@@

The results of running this simulation for 300 time steps are shown in
Fig.\,\ref{f:ch}. Fig.\,\ref{f:ch}A indicates the colour code used for retinal
cell position; dorsal retinal cells are indicated by green, nasal cells by
red. Lightness gives position along the relevant axis so that cells at the
dorsal-nasal corner are yellow and those at the ventral-temporal corner are
black. Fig.\,\ref{f:ch}E shows the expected layout of retinal ganglion cell
axons on the tectum. The axons retain the colour of their originating cell on
the retina. Nasal cell axons should be arranged
caudally; dorsal axons laterally; axons should be evenly spaced and the
pattern should span the entire tectum.  Fig.\,\ref{f:ch}, panels B--D show
`fish net' plots of the average location of the 4 branches of each axon (the
axon's centroid) at three timepoints. Fig.\,\ref{f:ch}B indicates the
randomised placement of axon branches at the rostral end of the tectum as
specified by the initial conditions (Eq.\,\ref{e:ic}). In these plots, lines are drawn between
axons that are expected to arrange themselves adjacently to one another by the
end of the simulation. The number of line crossings in this plot gives an
indication of the disorder in the arrangement. Fig.\,\ref{f:ch}G shows the positions of the individual branches
(i.e.~growth cones) at $t=300$ that contribute to the plot in
Fig.\,\ref{f:ch}D. By this time, the topological order matches that shown in
Fig.\,\ref{f:ch}E. Opposing gradients have sorted the axons along the two axes
to form a grid of cells. However, the arrangement does not span the full
tectum. Fig.\,\ref{f:ch}F shows two pattern metrics; $\epsilon$ is the mean
distance error of the axon centroids, given by:
%
\begin{equation}\label{e:eps}
\epsilon = \sqrt{\frac{\sum_k (\mathbf{x}_{k} - \mathbf{x}'_{k})^2}{n}},
\end{equation}
%
where $\mathbf{x}_{k}$ is the location of the centroid of the
branches of axon $k$ and $\mathbf{x}'_{k}$ the target tectal location for axon
$k$. $n$ is the number of RGC axons in the system.
%
$\epsilon$ tends to 0 as the pattern approaches the target layout. However,
two patterns may have the same non-zero value for $\epsilon$, despite one
being qualitatively more disordered than the other. For this reason, we also
use the number of crossings in the fish net plot of Fig.\,\ref{f:ch}D as a
second metric, called $\eta$, which tends towards an expected target number of
crossings, $\eta_t$ as the pattern approaches the target pattern.  For the wildtype
pattern, $\eta_t=0$, as in Fig.\,\ref{f:ch}E, although note that in some patterns,
$\eta_t$ is non-zero. Fig.\,\ref{f:ch}H shows selected
axons (one from the centre and one from each corner of the retina) and their
position history.

% ./build/sim/agent/agent1 configs/simpler/m_ee_G.json configs/simpler/e_wt_fig2.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_G_wt_fig2.png}
\caption{The pure gradient-following model. \textbf{A} The origin position
on the tectum gives axons a colour according to a dual-colour map. Axons in
panels B-H that are red are those which originated in the most naso-ventral
corner of the square simulated retina. \textbf{B}-\textbf{D} Axon centroid
positions on the simulated tectum (dashed line) at time 0, 40 and 300
units. Lines show adjacency relationships between the axons' origin positions
on the retina. \textbf{E} Layout of axon centroids on the tectum observed in
experimental results. \textbf{F} The sum-of-squared difference error,
$\epsilon$, and the number-of-crossings error metric, $\eta$ of the tectal axon
arrangement plotted with respect to simulation time. \textbf{G} The
arrangement of axon branches at t=300. \textbf{H} Selected axon locations at
t=300, with path histories in grey.  \emph{Abbreviations: D dorsal, V ventral, T
temporal, N nasal, C caudal, R rostral, M medial, L lateral.}}
\label{f:ch}
\end{figure}

The form of the retinal receptor expression (Eq.\,\ref{e:retrcpt}) is based on
empirical measurement, but the tectal ligand expression
(Eq.\,\ref{e:tecligexp}) was chosen arbitrarily. To expand the pattern in
Fig.\,\ref{f:ch}D so that it covers the entire tectum, we needed only to
adjust the tectal ligand expression, substituting the alternative exponential
expression:
%
\begin{equation} \label{e:tecligexp2}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(1.1 u)
\end{equation}
%
for Eq.\,\ref{e:tecligexp}.

The result at $t=300$ is shown in Figs.\,\ref{f:chalt}A
and \ref{f:chalt}B. The axon centroids are arranged across the full tectal
surface.  The arrangement is slightly disordered around the middle of the
tectum. This is caused by numerical artefacts from the discretization of the
ligand expression and the fact that in the centre, the difference between the
opposing receptor-ligand signals is very small. Nevertheless, the result
demonstrates that a modified form for the tectal ligand expression (which was
found easily by hand tuning) allows a well ordered map to form for the
wildtype system. Fig.\,\ref{f:chalt}C shows the error metric, $\epsilon$,
which is much closer to 0 with the new ligand expression \textbf{Fix graph}.

% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_wt_fig3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_eE_G_wt_fig3.png}
\caption{By tuning the tectal ligand expression (replacing Eq.\,\ref{e:tecligexp} with
 Eq.\,\ref{e:tecligexp2}), the pure gradient-following model can achieve a
 successful layout in which the axons are arranged evenly across the full
 tectum. At the centre of the tectum, where neither retinal receptor
 expression nor tectal ligand expression is high, the ordering is weaker than
 towards the edge. \textbf{A} Axon branches \textbf{B} axon
 centroids \textbf{C} Five selected axons.}
\label{f:chalt}
\end{figure}

@@

Fig\,\ref{f:chsurg} shows the results of 6 simulated surgical
manipulations. In A, a patch of tectal tissue has been cut out and rotated by
a quarter turn (the manipulation illustrated in Fig.\,\ref{f:trot90}). For
this simulation, the figure shows i) The locations of the axon centroids at
simulation end, ii) (inset) the expected final layout of the tectum, according
to the results of experimental work, with the colour code giving the origin
location of the axon's soma on the retina and iii) the error metric,
$\epsilon$ plotted versus simulation time. Simulation time was 1000 time
steps. The final locations of branches and axon centroids indicates that a
qualitative rotation of the graft patch has occurred, with red branches inside
the patch being located at the lateral-caudal corner rather than the
medial-caudal corner. A portion of the graft region (the medial side, after
rotation) is free of branches, indicating that the pattern is partial. A
similar result is shown for the 180\degree~example in B, in which the caudal
half of the rotated patch is free of branches, but the rostral half contains
branches with the correctly rotated orientations.

Fig.\,\ref{f:chsurg}C shows the result of the patch swap manipulation, in
which 12$\times$4 element grafts, one from the caudal side of the tectum and one from
the rostral side are swapped. This simulation is unsuccessful in patterning
the caudal half of the tectum. The large gradients, both positive and
negative, caused by the discontinuities along the long edges of the grafts
attracts almost all the branches (whose initial position is the rostral edge)
and very few pass through to the caudal side.

The retinal ablation shown in Fig.\,\ref{f:chsurg}D is also unsuccessful, with
the axons from the surviving nasal retina (the temporal half was removed)
failing to arrange themselves across the entire tectum, due to the pure
`position specification' which the dual gradient model achieves. Here, the
average axon position on the rostral-caudal axis of the tectum is plotted versus the average
nasal-temporal origin of the axon on the retina. Similarly,
the tectal ablation simulation fails to reproduce experimentally observed
results in which the retinal map `squashes up' within the remaining tectal
region. Instead, axons from the temporal retina arrange with normal spacing
and axons from the nasal retina are pushed to the caudal-most side of the
remaining tectum (Fig.\,\ref{f:chsurg}E). A similar result occurs in the
mismatch experiment (Fig.\,\ref{f:chsurg}F), but in this case there are no
axons from the temporal retina, which was been ablated and only the axons
from the nasal retina are present, arranged in a line.

% The chemo manipulations figure is made up from 4 figures generated with
% graph_layout::g
%
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecrot90_fig4row1.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecrot180_fig4row2.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecswap_fig4row3.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_retablat_fig4row4.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecablat_fig4row5.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_mismatch_fig4row6.json
%
% (see scripts/fig_G_surgical.sh)
\begin{figure}
\includegraphics[width=0.95\linewidth]{./images/fig_chemo_manipulations.png}
\caption{Surgical manipulations in the chemoaffinity-only model. \textbf{A} 90 degree
rotation. \textbf{B} 180 degree rotation. \textbf{C} Graft swap. \textbf{D}
Retinal ablation \textbf{E} Tectal ablation \textbf{F} mismatch ablation}
\label{f:chsurg}
\end{figure}

@@

\citet{simpson_simple_2011} showed that repulsive interactions combined with a
chemoaffinity-like effect could account for the surgical manipulations shown
in Figs.\,\ref{f:trot90}--\ref{f:chsurg}. To confirm that a competitive
mechanism could operate successfully in combination with the
gradient-following chemoaffinity mechanism described above, we introduced
three competition mechanisms and investigated their behaviour.

@@

Space based competition was implemented as a repulsive effect acting between
any two branches, whether they originated from different retinal ganglion
cells or from the same cell. An assumption is that there is some kind of
self-other signal which acts to discourage two cells from approaching each
other. A competition movement vector, $\mathbf{C}_b$, acting on branch $b$, is
defined as a proximity-weighted sum of the repulsion due to all nearby branches:
%
\begin{equation}
\mathbf{C}_b = \frac{1}{|B_{b,c}|} \sum_k \hat{\mathbf{x}}_{kb} W_c
\end{equation}
where $\hat{\mathbf{x}}_{kb}$ is the unit vector from branch $k$ to branch
$b$. The distance-dependent weight $W_c$ is given by
\begin{equation}
W_c(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_c}   & d_{kb} \leq 2r_c \\
     0 & d_{kb} > 2r_c
     \end{cases}
\end{equation}
for two growth cones of interaction radius $r_c$ separated by a distance
$d_{kb}$. $B_{b,c}$ is the set of growth cones within a distance $2r_c$ of
branch $b$. This description of the competition matches that
in \citet{simpson_simple_2011}.

@@

The model defined by the receptor and ligand expressions of Eqs.\,\ref{e:retrcpt}
\& \ref{e:tecligexp} along with Eq.\,\ref{e:mv2} with $m_{\!_I}=m_{\!_J}=0$ has three
free parameters; $m_{\!_G}$, $m_c$ and $r_c$. To find the best combination of
these parameters, we reasoned that evolution would optimize the arrangement of
the wildtype pattern. We performed a simulated annealing optimization to
minimize the error metric, $\epsilon$, for the wildtype pattern
(Fig.\,\ref{f:ch}E) at simulation time 1000. The parameters returned were
$m_{\!_G}$ = 0.003841, $m_c$ = 0.09959 and $r_c$ = 0.39918. Fig.\,\ref{f:GC}
shows the $\mathbf{GC}$ simulation with these parameters.

% ./scripts/fig_GC_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GC_surgical.png}
\caption{$\mathbf{GC}$ model, surgical manipulations}
\label{f:GCsurg}
\end{figure}

@@

% ./build/sim/agent/agent1 configs/simpler/m_ee_GI_best_1.json configs/simpler/e_wt_figcomp2.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GI_best_1_wt_figcomp2.png}
\caption{$\mathbf{GI}$ model}
\label{f:GI}
\end{figure}

@@

@@

>>>>

% Use the article doc class, with an 11 pt basic font size
\documentclass[11pt, a4paper, draft]{article}
%\documentclass[11pt, a4paper]{article}

@@

Each retinal ganglion cell (RGC) projects growth cones (referred to
as \emph{branches}) which carry a set of $N$ receptors, $r_i$ indexed by $i$ and
expressed at levels determined by the cell soma's location on the
retinal grid. ($N$ ligands, $l_i$, are also expressed by RGCs, but these are not
used by the chemoaffinity simulation.) The expression of each receptor
varies only with respect to one spatial dimension.

The minimum number of gradients required to provide a coordinate system on the
surface of the tectum within which axons could find a target location is
$N=2$, arranged in orthogonal directions. However if there are only two
gradients, imagining that one guides axons laterally and the other guides
axons in the caudal direction, then it is necessary for there to be an
additional `stopping mechanism' to prevent all of the axons eventually being
guided to the most lateral/caudal position. To avoid the need for a different
stopping mechanism, we chose to model \emph{opposing pairs of orthogonal
gradients} with $N=4$. With opposing gradients, an axon which interacts most
strongly with a gradient in the caudal direction would be drawn to the caudal
side of the tectum, whereas an axon whose strongest interaction was to climb a
gradient in the rostral direction would be drawn to the rostral side. There is
evidence that multiple gene expression gradients are found in the brain, with
examples of opposing pairs including XX and YY.

@@

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression
as \citet{simpson_simple_2011} which is
\begin{equation} \label{e:retrcpt}
r_i(x,y) = 1.05 + 0.26 \exp(2.3 u)
\end{equation}
Where $u$ gives the direction with which the expression increases and is
substituted by $(1-x)$ for $r_0$, $(1-y)$ for $r_1$, $x$ for $r_2$ and by $y$ for
$r_3$ (Fig.\,\ref{f:ex}B).

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in
orthogonal pairs of gradients. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for the form of tectal ligand expression is more ambiguous than that for
retinal receptor expression (Fig.\,\ref{f:expressions} summarises results from
the literature). We kept in mind the
possibility that tectal ligand expression may be modelled by exponential,
linear or logarithmic functions, but initially, we set it to the same
exponential function used for retinal expression,
\begin{equation} \label{e:tecligexp}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(2.3 u),
\end{equation}
and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in place of $u$ to obtain
the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$
(Fig.\,\ref{f:ex}D).

@@

\color{black}
Each discrete, square element in Fig.\,\ref{f:ex}B projects one RGC axon with $M=4$
branches to the tectum, carrying the receptor expression at the retinal
element's location.
At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the branch movement vector, is the sum of a chemoaffinity
effect and a `border effect', $\mathbf{B}_b$, which acts to retain branches
within the tectal region:
%
\begin{equation} \label{e:mv}
\mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + 0.5 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$.

@@

The results of running this simulation for 300 time steps are shown in
Fig.\,\ref{f:G}. Fig.\,\ref{f:G}A indicates the colour code used for retinal
cell position; dorsal retinal cells are indicated by green, nasal cells by
red. Lightness gives position along the relevant axis so that cells at the
dorsal-nasal corner are yellow and those at the ventral-temporal corner are
black. Fig.\,\ref{f:G}E shows the expected layout of retinal ganglion cell
axons on the tectum. The axons retain the colour of their originating cell on
the retina. Nasal cell axons should be arranged
caudally; dorsal axons laterally; axons should be evenly spaced and the
pattern should span the entire tectum.  Fig.\,\ref{f:G}, panels B--D show
`fish net' plots of the average location of the 4 branches of each axon (the
axon's centroid) at three timepoints. Fig.\,\ref{f:G}B indicates the
randomised placement of axon branches at the rostral end of the tectum as
specified by the initial conditions (Eq.\,\ref{e:ic}). In these plots, lines are drawn between
axons that are expected to arrange themselves adjacently to one another by the
end of the simulation. The number of line crossings in this plot gives an
indication of the disorder in the arrangement. Fig.\,\ref{f:G}G shows the positions of the individual branches
(i.e.~growth cones) at $t=300$ that contribute to the plot in
Fig.\,\ref{f:G}D. By this time, the topological order matches that shown in
Fig.\,\ref{f:G}E. Opposing gradients have sorted the axons along the two axes
to form a grid of cells. However, the arrangement does not span the full
tectum. Fig.\,\ref{f:G}F shows two pattern metrics; $\epsilon$ is the mean
distance error of the axon centroids, given by:
%
\begin{equation}\label{e:eps}
\epsilon = \sqrt{\frac{\sum_k^n (\mathbf{x}_{k} - \mathbf{x}'_{k})^2}{n}},
\end{equation}
%
where $\mathbf{x}_{k}$ is the location of the centroid of the
branches of axon $k$ and $\mathbf{x}'_{k}$ the target tectal location for axon
$k$. $n$ is the number of RGC axons in the system.
%
$\epsilon$ tends to 0 as the pattern approaches the target layout. However,
two patterns may have the same non-zero value for $\epsilon$, despite one
being qualitatively more disordered than the other. For this reason, we also
use the number of crossings in the fish net plot of Fig.\,\ref{f:G}D as a
second metric, called $\eta$, which tends towards an expected target number of
crossings, $\eta_t$ as the pattern approaches the target pattern.  For the wildtype
pattern, $\eta_t=0$, as in Fig.\,\ref{f:G}E, although note that in some patterns,
$\eta_t$ is non-zero. Fig.\,\ref{f:G}H shows selected
axons (one from the centre and one from each corner of the retina) and their
position history.

% ./build/sim/agent/agent1 configs/simpler/m_ee_G.json configs/simpler/e_wt_fig2.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_G_wt_fig2.png}
\caption{The pure gradient-following model. \textbf{A} The origin position
on the tectum gives axons a colour according to a dual-colour map. Axons in
panels B-H that are red are those which originated in the most naso-ventral
corner of the square simulated retina. \textbf{B}-\textbf{D} Axon centroid
positions on the simulated tectum (dashed line) at time 0, 40 and 300
units. Lines show adjacency relationships between the axons' origin positions
on the retina. \textbf{E} Layout of axon centroids on the tectum observed in
experimental results. \textbf{F} The sum-of-squared difference error,
$\epsilon$, and the number-of-crossings error metric, $\eta$ of the tectal axon
arrangement plotted with respect to simulation time. \textbf{G} The
arrangement of axon branches at t=300. \textbf{H} Selected axon locations at
t=300, with path histories in grey.  \emph{Abbreviations: D dorsal, V ventral, T
temporal, N nasal, C caudal, R rostral, M medial, L lateral.}}
\label{f:G}
\end{figure}

The form of the retinal receptor expression (Eq.\,\ref{e:retrcpt}) is based on
empirical measurement, but the tectal ligand expression
(Eq.\,\ref{e:tecligexp}) was chosen arbitrarily. To expand the pattern in
Fig.\,\ref{f:G}D so that it covers the entire tectum, we needed only to
adjust the tectal ligand expression, substituting the alternative exponential
expression:
%
\begin{equation} \label{e:tecligexp2}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(1.1 u)
\end{equation}
%
for Eq.\,\ref{e:tecligexp}.

The result at $t=300$ is shown in Figs.\,\ref{f:G0}A
and \ref{f:G0}B. The axon centroids are arranged across the full tectal
surface.  The arrangement is slightly disordered around the middle of the
tectum. This is caused by numerical artefacts from the discretization of the
ligand expression and the fact that in the centre, the difference between the
opposing receptor-ligand signals is very small. Nevertheless, the result
demonstrates that a modified form for the tectal ligand expression (which was
found easily by hand tuning) allows a well ordered map to form for the
wildtype system. Fig.\,\ref{f:G0}C shows the error metric, $\epsilon$,
which is much closer to 0 with the new ligand expression.

% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_wt_fig3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_eE_G_wt_fig3.png}
\caption{By tuning the tectal ligand expression (replacing Eq.\,\ref{e:tecligexp} with
 Eq.\,\ref{e:tecligexp2}), the pure gradient-following model can achieve a
 successful layout in which the axons are arranged evenly across the full
 tectum. At the centre of the tectum, where neither retinal receptor
 expression nor tectal ligand expression is high, the ordering is weaker than
 towards the edge. \textbf{A} Axon branches \textbf{B} axon
 centroids \textbf{C} Five selected axons.}
\label{f:G0}
\end{figure}

@@

Fig\,\ref{f:Gsurg} shows the results of 6 simulated surgical
manipulations. In A, a patch of tectal tissue has been cut out and rotated by
a quarter turn (the manipulation illustrated in Fig.\,\ref{f:trot90}). For
this simulation, the figure shows i) The locations of the axon centroids at
simulation end, ii) (inset) the expected final layout of the tectum, according
to the results of experimental work, with the colour code giving the origin
location of the axon's soma on the retina and iii) the error metric,
$\epsilon$ plotted versus simulation time. Simulation time was 1000 time
steps. The final locations of branches and axon centroids indicates that a
qualitative rotation of the graft patch has occurred, with red branches inside
the patch being located at the lateral-caudal corner rather than the
medial-caudal corner. A portion of the graft region (the medial side, after
rotation) is free of branches, indicating that the pattern is partial. A
similar result is shown for the 180\degree~example in B, in which the caudal
half of the rotated patch is free of branches, but the rostral half contains
branches with the correctly rotated orientations.

Fig.\,\ref{f:Gsurg}C shows the result of the patch swap manipulation, in
which 12$\times$4 element grafts, one from the caudal side of the tectum and one from
the rostral side are swapped. This simulation is unsuccessful in patterning
the caudal half of the tectum. The large gradients, both positive and
negative, caused by the discontinuities along the long edges of the grafts
attracts almost all the branches (whose initial position is the rostral edge)
and very few pass through to the caudal side.

The retinal ablation shown in Fig.\,\ref{f:Gsurg}D is also unsuccessful, with
the axons from the surviving nasal retina (the temporal half was removed)
failing to arrange themselves across the entire tectum, due to the pure
`position specification' which the dual gradient model achieves. Here, the
average axon position on the rostral-caudal axis of the tectum is plotted versus the average
nasal-temporal origin of the axon on the retina. Similarly,
the tectal ablation simulation fails to reproduce experimentally observed
results in which the retinal map `squashes up' within the remaining tectal
region. Instead, axons from the temporal retina arrange with normal spacing
and axons from the nasal retina are pushed to the caudal-most side of the
remaining tectum (Fig.\,\ref{f:Gsurg}E). A similar result occurs in the
mismatch experiment (Fig.\,\ref{f:Gsurg}F), but in this case there are no
axons from the temporal retina, which was been ablated and only the axons
from the nasal retina are present, arranged in a line.

% The chemo manipulations figure is made up from 4 figures generated with
% graph_layout::g
%
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecrot90_fig4row1.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecrot180_fig4row2.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecswap_fig4row3.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_retablat_fig4row4.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecablat_fig4row5.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_mismatch_fig4row6.json
%
% (see scripts/fig_G_surgical.sh)
\begin{figure}
\includegraphics[width=0.95\linewidth]{./images/fig_chemo_manipulations.png}
\caption{Surgical manipulations in the chemoaffinity-only model. \textbf{A} 90 degree
rotation. \textbf{B} 180 degree rotation. \textbf{C} Graft swap. \textbf{D}
Retinal ablation \textbf{E} Tectal ablation \textbf{F} mismatch ablation}
\label{f:Gsurg}
\end{figure}

@@

\citet{simpson_simple_2011} showed that repulsive interactions combined with a
chemoaffinity-like effect could account for the surgical manipulations shown
in Figs.\,\ref{f:trot90}--\ref{f:Gsurg}. To confirm that a competitive
mechanism could operate successfully in combination with the
gradient-following chemoaffinity mechanism described above, we introduced
three competition mechanisms and investigated their behaviour.

@@

Space based competition was implemented as a repulsive effect acting between
any two branches, whether they originated from different retinal ganglion
cells or from the same cell. An assumption is that there is some kind of
self-other signal which acts to discourage two cells from approaching each
other. A competition movement vector, $\mathbf{C}_b$, acting on branch $b$, is
defined as a proximity-weighted sum of the repulsion due to all nearby branches:
%
\begin{equation} \label{e:C}
\mathbf{C}_b = \frac{1}{|B_{b,c}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb} W_c
\end{equation}
where $\hat{\mathbf{x}}_{kb}$ is the unit vector from branch $k$ to branch
$b$. The distance-dependent weight $W_c$ is given by
\begin{equation}
W_c(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_c}   & d_{kb} \leq 2r_c \\
     0 & d_{kb} > 2r_c
     \end{cases}
\end{equation}
for two growth cones of interaction radius $r_c$ separated by a distance
$d_{kb}$. $B_{b,c}$ is the set of growth cones within a distance $2r_c$ of
branch $b$ and the total number of branches in the system is the number of RGC
axons, $n$ multiplied by the number of branches per axon, $M$. This
description of the competition matches that in \citet{simpson_simple_2011}.

@@

The model defined by the receptor and ligand expressions of Eqs.\,\ref{e:retrcpt}
\& \ref{e:tecligexp} along with Eq.\,\ref{e:mv2} with $m_{\!_I}=m_{\!_J}=0$ has three
free parameters; $m_{\!_G}$, $m_c$ and $r_c$. To find the best combination of
these parameters, we reasoned that evolution would optimize the arrangement of
the wildtype pattern. We performed a simulated annealing optimization to
minimize the error metric, $\epsilon$, for the wildtype pattern
(Fig.\,\ref{f:G}E) at simulation time 1000. The parameters returned were
$m_{\!_G}$ = 0.003841, $m_c$ = 0.09959 and $r_c$ = 0.39918. Fig.\,\ref{f:GC}
shows the $\mathbf{GC}$ simulation with these parameters. Fish net plots at
$t=40$ and $t=1000$ show that the pattern develops quickly and compares well
with the expected layout (Fig.\,\ref{f:G}E). We investigated the response of
this model to the same set of surgical manipulations shown for the
chemoaffinity-only model in Fig.\,\ref{f:Gsurg}. We found that the addition of
competition leads to a better correspondence to the experimental results than
for the chemoaffinity-only model (see supplementary Fig.\,\ref{f:GCsurg}).

Because the competition effect, $\mathbf{C}_b$ (Eq.\,\ref{e:C}), requires a
global supervisor to provide information to each branch about the location of
all other branches, we looked for a receptor-ligand interaction that could
signal between axon branches and enable them to adjust their movements away
from one another. We investigated the idea that the same receptor-ligand
interaction that is modelled between RGC axons and the tectum could also act
as a signal between the axons, because retinal ganglion cells are known to
express ephrin receptor ligands as well as Eph
receptors \citep{ephrins_on_RGCs}. We defined $N=4$ ligands, $l_i$, expressed
by the RGCs in the model according to
%
\begin{equation} \label{e:retlig}
l_i(x,y) = 1.05 + 0.26 \exp(2.3 u).
\end{equation}
%
We used exactly the same form as the receptor expression given by
Eq.\,\ref{e:retrcpt}, but with opposite gradient direction so here $u$ is
substituted by $x$ for $l_0$, $y$ for $l_1$, $(1-x)$ for $l_2$ and by $(1-y)$
for $l_3$ (see Fig.\,\ref{f:ex}C). We modelled a
multiplicative (or mass-action?) interaction to give the movement
vector, $\mathbf{J}_b$, as  follows:
%
\begin{equation}
% No \frac{1}{|B_b|} in this one. Really?
\mathbf{J}_b = \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,Q_{\!_J}(d_{kb})
\end{equation}
%
where $Q_{\!_J}(d_{kb})$ is the signalling strength between two growth cones of
receptor-ligand interaction radius $r_{\!_J}$ a distance $d_{kb}$ from one another,
given by
%
\begin{equation}
Q_{\!_J}(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r_{\!_J} \\
     0 & d_{kb} > 2r_{\!_J}
     \end{cases}
\end{equation}
%
The sign of $Q_{\!_J}$ (which is dependent on the values of $F_i$) determines whether
the interaction, $\mathbf{J}_b$, is repulsive or attractive. $l_{i,k}$ is the
expression of ligand $i$ on branch $k$.

Considering just one receptor-ligand pair,
for instance $r_0$/$l_0$, and defining their interaction as repulsive, an axon
from the transverse side of the retina, which expresses a high level of
receptor $r_0$ will interact most strongly with an axon from the nasal side of
the retina, where $l_0$ is most highly expressed. The interaction with axons
that originate nearby, on the transverse side will be minimal, as these
express low levels of $l_0$. Axons originating from the nasal side of the
retina interact weakly with axons originating nearby.

@@

% ./build/sim/agent/agent1 configs/simpler/m_ee_GI_best_1.json configs/simpler/e_wt_figcomp2.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GI_best_1_wt_figcomp2_exit_true.png}
\caption{$\mathbf{GI}$ model}
\label{f:GI}
\end{figure}

@@

\subsection*{Attractive receptor-ligand interactions}

If we switch a receptor-ligand pair to have an attractive interaction, the
tectal ligand expression should change from exponential to logarithmic.

@@

% ./scripts/fig_GC_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GC_surgical.png}
\caption{$\mathbf{GC}$ model, surgical manipulations. Demonstrating that this
model reproduces these manipulations.}
\label{f:GCsurg}
\end{figure}
